Bootstrap learning via modular concept discovery

### 2013-03-19 drafts/ijcai2013/ijcai13.tex
IJCAI 2013 submission changes before final submission.

\setlength{\droptitle}{0em}   
\date{}

@@

\maketitle
\begin{abstract}
 Suppose a learner is faced with a domain of problems about which it
 knows nearly nothing. It does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer it effectively. How can such a learner ever get off
 the ground? A common intuition is that if the solutions to these
 problems share a common structure, and the learner can solve some
 simple problems by brute force, it should be able to extract useful
 components from these solutions and, by composing them, explore the
 solution space more efficiently. Here, we formalize this intuition,
 where the solution space is the space of typed functional programs
 and the gained information is stored as a stochastic grammar over
 programs. We propose an iterative procedure for exploring such
 spaces: in the first step of each iteration, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. We test this
 procedure on symbolic regression and boolean circuit learning and
 show that the learner discovers modular concepts for these
 domains. Whereas the learner is able to solve almost none of the
 posed problems in the procedure's first iteration, it rapidly becomes
 able to solve a large number by gaining abstract knowledge of the
 structure of the solution space.

@@

But thes problems we want AI systems to solve often lack this kind of
built in structure. These are the kinds of problems more often
associated with ILP and classical AI than with GP and machine
learning. 
\\

@@

The remainder of this paper will include two major sections. In the
first, we describe in detail the EC algorithm and design choices we
make in implementing it. We will then present results from two
experiments in the domains of symbolic regression and Boolean
functioin learning.

@@

We propose the \textbf{E-C algorithm}, an iterative procedure to solve this
problem. It will be useful in presenting the algorithm to define the
following terminology: a \textbf{frontier} is the finite set of
expressions that the algorithm considers in any one iteration. The
\textbf{frontier size} $N$ is the prespecified number of elements in
the frontier. We will say that a task is \textbf{hit} by the frontier
if there is a expression in the frontier that solves it.

The E-C algorithm maintains a distribution $\mathcal{D}$ over
expressions in $\mathcal{L}$. At each iteration, it
\begin{enumerate}
\item \textbf{explores} the frontier -- enumerating the $N$ most probable expressions
  from the current distribution $\mathcal{D}$ -- and
\item \textbf{compresses} the hit tasks in the frontier to estimate a new
  distribution.
\end{enumerate}

Next, we specify a representation language $\mathcal{L}$, a
form for the distribution $\mathcal{D}$, and a compression criterion for a
set of expressions in $\mathcal{L}$. We also describe
procedures for the enumeration of the frontier, compression of the
expressions, and re-estimation of $\mathcal{D}$.

@@

Following~\cite{liang10programs} and~\cite{Briggs:2008}, we use a
simply typed combinatory logic -- a variable-free subset of the
polymorphic simply-typed lambda calculus~\cite{Pierce_2002} -- as our
program representation language. In short, the combinatory logic
introduces a \emph{basis} of several primitive \emph{combinators} such
that any function can be written as applications of the basis
combinators. Some common basis combinators are defined as follows:
\begin{align}
\mathbf{I}\; x &\rightarrow  x & \text{ (identity) }\\
\mathbf{S}\; f\; g\; x\; &\rightarrow (f\; x)\; (g\; x)\\ 
\mathbf{C}\; f\; g\; x\; &\rightarrow (f\; x)\; g  \\ 
\mathbf{B}\; f\; g\; x\; &\rightarrow f\; (g\; x) & \text{ (composition) }
\end{align}

@@

The basis combinators can themselves be expressed in the lambda
calculus (with definitions following directly from the equations
above). The lambda calculus has two basic operations -- application
and abstraction (using the $\lambda$ operator -- but in using the
combinatory logic we sequester uses of the abstraction operation
inside the combinators. In doing so, we have replaced the variable
binding of the $\lambda$ operator with the variable \emph{routing} of
these basis combinators. Our representation thus becomes
variable-free. See Liang et al.~\cite{liang10programs} for a more
detailed discussion of this routing interpretation.

@@

We specify a distribution $\mathcal{D}$ over expressions in
$\mathcal{L}$ as a stochastic
grammar~\cite{feldman1969grammatical}. Many stochastic grammars have
the desirable property that the probability of an expression is the
product of the probabilities of its subcomponents. The distribution we
specify will be a simple verson of this, in which the
probability of an expression is the product of the probability of its
leaves.

@@

Now suppose that we have a set of expressions $\mathcal{E}$, and we
wish to estimate the maximum likelihood values of the $p_n$'s. If the
choice of primitive at each leaf were not conditioned on the
requesting type, we could just count all occurences of each primitive
in our expression set, and this would be proportional to the maximum
likelihood estimate. This is the ML estimator for a multinomial
distribution. However, the fact that we condition on the requesting
type makes this a considerably more difficult optimization task. One
of the simplest approximations we can make is to calculate the
frequency with which each primitive combinator appears in
$\mathcal{E}$ when the requesting type is such that it could have
been chosen. This is a straightforward calculuation, and
we use it for its convenience, finding that our empirical results
justify its use. Future inquiry will be needed to determine to what
extent a more accurate estimator can improve these results.

\subsection{Best-first enumeration of programs}
In order to explore the frontier of most promising programs, we need a
procedure that enumerates the $N$ most probable programs. There has
been recent interest in this problem, most of it focused on
enumerating the shortest program satisfying some
criterion~\cite{DBLP:conf/sfp/Katayama05}
\cite{DBLP:conf/aaip/YakushevJ09}. Our approach is to formulate the
problem as a best-first exploration of an AND/OR tree
(Figure~\ref{fig:andor})~\cite{nilsson1982principles}~\cite{DBLP:journals/cacm/Hall73}. In
words, the enumeration procedure is best described by the following
recursion: every program is either a primitive combinator or it is a
left child program applied to a right child program.

@@

\setcounter{subfigure}{0} 
\begin{figure}[ht]
\centering
\begin{subfigure}[Before]{\linewidth}
\includegraphics[width=.9\linewidth]{figures/symreg3frontiers.pdf}
\caption{Learning curves as a function of frontier size.  As frontier
  size is increased, curves plateau closer to $100\%$ performance. A
  baseline search over $150000$ expressions only hits $3\%$ of the
  tasks. \label{fig:symregLearning}}
\end{subfigure}

\begin{subfigure}[After]{\linewidth}
\includegraphics[width=.9\linewidth]{figures/symRegDiffCurricula.pdf}
\caption{How do different task sets affect learning curves? Learning
  curves at frontier size of $10000$ for different task
  sets. Performance is equally good if either the constant functions
  of the linear functions are missing, but drops dramatically when both
  are absent, even though these only account for about $10\%$ of the
  tasks. Performance drops even lower if we only consider ``complex''
  quadratics.}
\label{fig:symregCurricula}
\end{subfigure}
%% \begin{subfigure}[After]{.3\linewidth}\
%% \includegraphics[width=.9\linewidth]{figures/symregCircuitDepths.pdf}
%% \caption{Circuit depth as a function of iteration on a typical run.}
%% \label{fig:symregDepths}
%% \end{subfigure}
\vspace{-.2cm}
\caption{Symbolic regression experiment results.}
\label{fig:symreg} 
\end{figure}

@@

What does the E-C algorithm learn in this task?  Inspecting the top
weighted primitives of the final grammars, we find many incrementers
(e.g. (+1), (+2), etc.), several versions of expressions that
translate to functions like $x*f(x)$ and $f(x*f(x))$, and combinations
of these, like $x*(x+3)+3$ and $x*(x+2)$. Informally, we see
expressions that apply functions to themselves, building up complex
functions with relatively few unique primitives.

To what degree is ``conceptual bootstrapping'' responsible for the
improved performance. That is, to what extent does the presence of
simple problems account for the ability to learn complex functions?
One hypothesis is that to succeed on a set of tasks, the set must
contain a ``learnable'' curriculum, a set of tasks that serve as
stepping stones from an impoverished representation to a rich one. We
can test this hypothesis by varying the set of tasks to which the EC
algorithm is exposed. If it is true, then we should see a nonlinear
reponse to reducing the number of easy tasks, as the curriculum
changes from being learnable to being unlearnable.

@@

\begin{figure}[ht]
\includegraphics{./tablealonedoc.pdf}
\vspace{-.5cm}
\caption{Some learned primitives from the circuit based Boolean
  function learning experiment. We highlight compression using $NOT$,
  $E_1$ and $E_2$. \,(a)The three elementary gates used to generate
  the task set. (b) Two higher-order primitives in the induced
  grammer. Note how $E_1$ can be used to define FALSE by taking the
  NOT function as an argument. (c) Additional common Boolean functions
  found in the learned grammar.  \label{table:booleanCircuits}}
\vspace{-.4cm}
\end{figure}

@@

\begin{figure}
\vspace{-.3cm}
\includegraphics[width=\linewidth]{figures/circuitDistrComp.pdf}
\vspace{-.6cm}
\caption{Distribution of boolean functions among (a) the set of 1000
  circuits randomly constructed from AND, OR and NOT gates (see text);
  b) the first 1000 expressions from the grammar over expressions
  \emph{before} learning; c) the first 1000 expressions from the
  grammar over expressions \emph{after} learning. Gray levels indicate
  percent of instances (circuits/expressions) which evaluate to that
  function.\label{fig:booldistr}}
\end{figure}

@@

\begin{figure}[h]
\begin{subfigure}{.1\linewidth}
\caption{\label{fig:boolLearningCurvesCircuits} }
\end{subfigure}
\begin{subfigure}[Before]{.9\linewidth}
\includegraphics[width=.9\linewidth]{figures/booleancircuits3frontiers.pdf}
\end{subfigure}

\vspace{-.2cm}
\begin{subfigure}{.1\linewidth}
\caption{\label{fig:boolLearningCurvesUniform}}
\end{subfigure}
\begin{subfigure}[After]{.9\linewidth}
\includegraphics[width=.9\linewidth]{figures/booleanfunctionsUniform.pdf}

@@

Our specific implementation of this approach involved a number of
choices which should be explored further in future work.  Our claim
here is not that the EC algorithm is optimal, but that it captures key
features of how people solve the problem of learning new systems of
concepts in acquiring deep domain expertise.  Much like a human
researcher investigating a new area of study, our algorithm tries
different models, watches them fail or succeed, and abstracts out the
model elements that capture relevant differences.  It recombines and
reuses succcessful patterns of reasoning that span many problems in a
domain.

At their best, human processes of discovery also clearly go beyond
what our algorithm is capable of.  The algorithm can get stuck in
plateaus after a rapid initial increase in performance, due to its
circular nature: it only inspects the region of the solution space
that is typical of solutions already found.  Humans can become stuck
in similar ways, but at least sometimes they can realize that they are
stuck, and attempt to find another class of concepts for the remaining
unsolved tasks.  Our algorithm requires, as human learners do, a
sufficiently graded ``curriculum'', or spectrum of problems to solve --
simple problems provide necessary stepping stones to building complex
representations.  Also like humans, our algorithm can adaptively
``self-pace'' its ways through the curriculum, figuring out which
problems it is ready to tackle when, which problems are appropriate in
difficulty given what it currently knows.  Yet people can sometimes
more intelligently compose and modify the set of problems under
consideration, designing their own stepping stones to expand the set
of problems they can solve.  Making bootstrap learning systems smarter
in these more human-like ways is a prime target for future research.
%% \subsection{Digital arithmetic}
%% Having demonstrated the behavior of the E-C algorithm on symbolic
%% regression, we analyze its performance on tasks involving digital
%% arithmetic, a class of problems with known modular structure. The
%% general goal is to discover boolean circuits consisting of a limited
%% set of gates that perform arithmetic operations on binary number
%% representations.

>>>>

\setlength{\droptitle}{-1em}   
\date{}

@@

\maketitle
\begin{abstract}
%\small
 Suppose a learner is faced with a domain of problems about which it
 knows nearly nothing. It does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer it effectively. How can such a learner ever get off
 the ground? A common intuition is that if the solutions to these
 problems share a common structure, and the learner can solve some
 simple problems by brute force, it should be able to extract useful
 components from these solutions and, by composing them, explore the
 solution space more efficiently. Here, we formalize this intuition,
 where the solution space is the space of typed functional programs
 and the gained information is stored as a stochastic grammar over
 programs. We propose an iterative procedure for exploring such
 spaces: in the first step of each iteration, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. We test this
 procedure on symbolic regression and boolean circuit learning and
 show that the learner discovers modular concepts for these
 domains. Whereas the learner is able to solve almost none of the
 posed problems in the procedure's first iteration, it rapidly becomes
 able to solve a large number by gaining abstract knowledge of the
 structure of the solution space.

@@

But, as mentioned before, the problems we want AI systems to solve
often lack this kind of built in structure. Here we attempt to learn
functional programs to solve those kinds of problems.  \\

@@

The remainder of this paper will include two major sections. In the
first, we describe in detail the EC algorithm and design choices we
make in implementing it. We will then present results from two
experiments in the domains of symbolic regression and Boolean
function learning.

@@

We propose the \textbf{EC algorithm}, an iterative procedure to solve this
problem. It will be useful in presenting the algorithm to define the
following terminology: a \textbf{frontier} is the finite set of
expressions that the algorithm considers in any one iteration. The
\textbf{frontier size} $N$ is the prespecified number of elements in
the frontier. We will say that a task is \textbf{hit} by the frontier
if there is a expression in the frontier that solves it.

The EC algorithm maintains a distribution $\mathcal{D}$ over
expressions in $\mathcal{L}$. At each iteration, it
\begin{enumerate}
\item \textbf{explores} the frontier -- enumerating the $N$ most probable expressions
  from the current distribution $\mathcal{D}$ -- and
\item \textbf{compresses} the hit tasks in the frontier to estimate a new
  distribution.
\end{enumerate}

@@

Following~\cite{liang10programs} and~\cite{briggs2006functional}, we use a
simply typed combinatory logic -- a variable-free subset of the
polymorphic simply-typed lambda calculus~\cite{Pierce_2002} -- as our
program representation language. In short, the combinatory logic
introduces a \emph{basis} of several primitive \emph{combinators} such
that any function can be written as applications of the basis
combinators. Some common basis combinators are defined as follows:
\begin{align}
\mathbf{I}\; x &\rightarrow  x & \text{ (identity) }\\
\mathbf{S}\; f\; g\; x\; &\rightarrow (f\; x)\; (g\; x)\\ 
\mathbf{C}\; f\; g\; x\; &\rightarrow (f\; x)\; g  \\ 
\mathbf{B}\; f\; g\; x\; &\rightarrow f\; (g\; x) & \text{ (composition) }
\end{align}

@@

The basis combinators can themselves be expressed in the lambda
calculus (with definitions following directly from the equations
above). The lambda calculus has two basic operations -- application
and abstraction -- but in using the
combinatory logic we sequester uses of the abstraction operation
inside the combinators. In doing so, we have replaced the variable
binding of the $\lambda$ operator with the variable \emph{routing} of
these basis combinators. Our representation thus becomes
variable-free. See Liang et al.~\cite{liang10programs} for a more
detailed discussion of this routing interpretation.

@@

We specify a distribution $\mathcal{D}$ over expressions in
$\mathcal{L}$ as a stochastic
grammar~\cite{feldman1969grammatical}. Many stochastic grammars have
the desirable property that the probability of an expression is the
product of the probabilities of its subcomponents. The distribution we
specify will be a simple version of this, in which the
probability of an expression is the product of the probability of its
leaves.

@@

Now suppose that we have a set of expressions $\mathcal{E}$, and we
wish to estimate the maximum likelihood values of the $p_n$'s. If the
choice of primitive at each leaf were not conditioned on the
requesting type, we could just count all occurrences of each primitive
in our expression set, and this would be proportional to the maximum
likelihood estimate. This is the ML estimator for a multinomial
distribution. However, the fact that we condition on the requesting
type makes this a considerably more difficult optimization task. One
of the simplest approximations we can make is to calculate the
frequency with which each primitive combinator appears in
$\mathcal{E}$ when the requesting type is such that it could have
been chosen. This is a straightforward calculation, and
we use it for its convenience, finding that our empirical results
justify its use. Future inquiry will be needed to determine to what
extent a more accurate estimator can improve these results.

\subsection{Best-first enumeration of programs}
In order to explore the frontier of most promising programs, we need a
procedure that enumerates the $N$ most probable expressions. There has
been recent interest in this problem, most of it focused on
enumerating the shortest program satisfying some
criterion~\cite{DBLP:conf/sfp/Katayama05}
\cite{DBLP:conf/aaip/YakushevJ09}. Our approach is to formulate the
problem as a best-first exploration of an AND/OR tree
(Figure~\ref{fig:andor})~\cite{nilsson1982principles}~\cite{DBLP:journals/cacm/Hall73}. In
words, the enumeration procedure is best described by the following
recursion: every program is either a primitive combinator or it is a
left child program applied to a right child program.

@@

\setcounter{subfigure}{0} 
\begin{figure}[ht]
\centering
\includegraphics[width=.8\linewidth]{figures/symreg3frontiers.pdf}
\vspace{-.2cm}
\caption{Learning curves as a function of frontier size.  As frontier
  size is increased, curves plateau closer to $100\%$ performance. A
  baseline search over $150000$ expressions only hits $3\%$ of the
  tasks. \label{fig:symregLearning}}
\vspace{-.3cm}
\end{figure}
\begin{figure}[ht]
\centering
\includegraphics[width=.8\linewidth]{figures/symRegDiffCurricula.pdf}
\caption{How do different task sets affect learning curves? Learning
  curves at frontier size of $10000$ for different task
  sets.} 
\label{fig:symregCurricula}
\vspace{-.5cm}
\end{figure}

@@

What does the EC algorithm learn in this task?  Inspecting the top
weighted primitives of the final grammars, we find many incrementers
(e.g. (+1), (+2), etc.), several versions of expressions that
translate to functions like $x*f(x)$ and $f(x*f(x))$, and combinations
of these, like $x*(x+3)+3$ and $x*(x+2)$. Informally, we see
expressions that apply functions to themselves, building up complex
functions with relatively few unique primitives.

To what degree is ``conceptual bootstrapping'' responsible for the
improved performance. That is, to what extent does the presence of
simple problems account for the ability to learn complex functions?
One hypothesis is that to succeed on a set of tasks, the set must
contain a ``learnable'' curriculum, a set of tasks that serve as
stepping stones from an impoverished representation to a rich one. We
can test this hypothesis by varying the set of tasks to which the EC
algorithm is exposed. If it is true, then we should see a nonlinear
response to reducing the number of easy tasks, as the curriculum
changes from being learnable to being unlearnable.

@@

\begin{figure}[t]
\includegraphics{./tablealonedoc.pdf}
\vspace{-.5cm}
\caption{Some learned primitives from the circuit based Boolean
  function learning experiment. We highlight compression using $NOT$,
  $E_1$ and $E_2$. \,(a)The three elementary gates used to generate
  the task set. (b) Two higher-order primitives in the induced
  grammar. Note how $E_1$ can be used to define FALSE by taking the
  NOT function as an argument. (c) Additional common Boolean functions
  found in the learned grammar.  \label{table:booleanCircuits}}
\vspace{-.4cm}
\end{figure}

@@

\begin{figure}
\vspace{-.3cm}
\includegraphics[width=\linewidth]{figures/circuitDistrComp.pdf}
\vspace{-.6cm}
\caption{Distribution of boolean functions among (a) the set of 1000
  circuits randomly constructed from AND, OR and NOT gates (see text);
  b) the first 1000 expressions from the grammar over expressions
  \emph{before} learning; c) the first 1000 expressions from the
  grammar over expressions \emph{after} learning. Gray levels indicate
  percent of instances (circuits/expressions) which evaluate to that
  function.\label{fig:booldistr}}
\vspace{-.4cm}
\end{figure}

@@

\vspace{-.3cm}
\begin{figure}[h]
\begin{subfigure}{.1\linewidth}
\caption{\label{fig:boolLearningCurvesCircuits} }
\end{subfigure}
\begin{subfigure}[Before]{.9\linewidth}
\includegraphics[width=.9\linewidth]{figures/booleancircuits3frontiers.pdf}
\end{subfigure}

\vspace{-.32cm}
\begin{subfigure}{.1\linewidth}
\caption{\label{fig:boolLearningCurvesUniform}}
\end{subfigure}
\begin{subfigure}[After]{.9\linewidth}
\includegraphics[width=.9\linewidth]{figures/booleanfunctionsUniform.pdf}

@@

Our specific implementation of this approach involved a number of
choices which should be explored further in future work.  Our claim
here is not that the EC algorithm is optimal, but that it captures key
features of how people solve the problem of learning new systems of
concepts in acquiring deep domain expertise.  Much like a human
researcher investigating a new area of study, our algorithm tries
different models, watches them fail or succeed, and abstracts out the
model elements that capture relevant differences.  It recombines and
reuses successful patterns of reasoning that span many problems in a
domain.

At their best, human processes of discovery clearly go beyond
what our algorithm is capable of.  The algorithm can get stuck in
plateaus after a rapid initial increase in performance, due to its
circular nature: it only inspects the region of the solution space
that is typical of solutions already found.  Humans can become stuck
in similar ways, but at least sometimes they realize that they are
stuck and attempt to find another class of concepts for the remaining
unsolved tasks.  Our algorithm requires, as human learners do, a
sufficiently graded ``curriculum'', or spectrum of problems to solve --
simple problems provide necessary stepping stones to building complex
representations.  Also like humans, our algorithm can adaptively
``self-pace'' its ways through the curriculum, figuring out which
problems it is ready to tackle when, which problems are appropriately
difficult given what it currently knows.  Yet people can sometimes
more intelligently compose and modify the set of problems under
consideration, designing their own stepping stones to expand the set
of problems they can solve.  Making bootstrap learning systems smarter
in these more human-like ways is a prime target for future research.
%% \subsection{Digital arithmetic}
%% Having demonstrated the behavior of the E-C algorithm on symbolic
%% regression, we analyze its performance on tasks involving digital
%% arithmetic, a class of problems with known modular structure. The
%% general goal is to discover boolean circuits consisting of a limited
%% set of gates that perform arithmetic operations on binary number
%% representations.
